---
title: "biochar_pca"
author: "Shahnewaz Jim Ahmad"
date: "2023-12-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up directory
```{r setting directory}
getwd()
setwd("C:/Users/shahn/Downloads/EnvEng/Winter_23/EnvDataSci/Assignments/Thesis")
``` 
  
## data and library
```{r data and library}
data = read.csv(file = "biochar_data_R.csv",
                header = T,
                sep = ";",
                quote = "", 
                dec = ","
                )
library(corrr)      #correlation analysis, mainly handle R dataframe
library(ggcorrplot) #easy to vis correlation matrix
library(FactoMineR) #for multivariate analysis, give access to the PCA module
library(tidyverse)  #
library(vip)        #plotting variable importance for prediction in a model  
library(caret)      #for cross validation
library(utils)      #for R documentation and package management
library(GGally)     #specially for ggpir package in this case to visulaize scatterplot and correlation coefficient as matrices
library(patchwork)


str(data)
#data = data[, c(1:3, 5, 9:10, 15, 18:19)]
#data %>% rmarkdown::paged_table()
#colSums(is.na(data))
num_data <- data[,c(2:10, 14:16, 18:19)]
num_data %>% rmarkdown::paged_table()

dt <- read.csv(file = "Agg&SOC_new.csv",
               header = TRUE,
               sep = ",",
               quote = "",
               dec = "."
               )
dt <- dt %>% 
  mutate(dt,
         category = rep(c("a", "b", "c", "d", "e", "f", "g", "h", "i", "j", "k", "l", "m", "n", "o"), each = 4)
         )
```
## Agg Fraction bar table

```{r}
mean_dt <- dt %>%
  group_by(Treatments, Agg_size) %>% 
  summarise(mean= mean(Fraction_.),
            se = sd(Fraction_.)/sqrt(n())
            )
#mean_dt <- mean_dt %>%
#  group_by(Treatments, Agg_size) %>% 
#  mutate(sig = rep(c("a", "ab", "a", "ab", "a", "ab", "a", "ab", "a", "ab", "a", "ab", "a", "ab", "a", "ab", "a", "ab", "a", "ab"), length.out = n()))

ggplot(mean_dt, aes(fill = Agg_size, y = mean, x = Treatments)) + 
  geom_bar(position = "dodge", stat = "identity")+
  geom_errorbar(aes(ymin = mean-se, ymax = mean+se),
                position =  position_dodge(width = 0.9), width = 0.3)+
  #scale_fill_brewer(palette = "BuPu")
  scale_fill_grey(start = 0.75, end = 0.25)+
  labs(x = "Treatments",
       y = "The Proportions of Aggregates %")

ggsave("agg.png")
```

#MWD

```{r}
sum_data <- dt %>% 
  
  group_by(Treatments, category) %>% 
  summarise(sum_WD = sum(WD))

mean_data <- sum_data %>% 
  group_by(Treatments) %>% 
  summarise(mean_WD = mean(sum_WD),
            se_WD = sd(sum_WD)/sqrt(n())
            )

mean_data <- mean_data %>% 
  mutate(mean_data, tuk = c("c","bc","b","ab","a") )

ggplot(mean_data, aes(x = Treatments, y = mean_WD, fill = Treatments))+
  geom_bar(stat = "identity",  color="black", fill="white")+
  #scale_fill_grey(start = 0.75, end = 0.75)
  geom_errorbar(aes(ymin = mean_WD-se_WD, ymax = mean_WD+se_WD),
                position =  position_dodge(width = 0.9), width = 0.3)+
  labs(x = "Treatments",
       y = "Mean Weight Daiameter (mm)")+
  geom_text(label = mean_data$tuk, nudge_y = 0.098, size = 5)
  #geom_text(aes(label=Tukey))
ggsave("mwd.png")

sum_data_imp <- dt %>% 
  
  group_by(Treatments, Agg_size) %>% 
  summarise(sum_WD_imp =sum(WD))


ggplot(sum_data_imp, aes(x = Treatments, y = sum_WD_imp, fill = Agg_size))+
  geom_bar(stat = "identity", position = "stack")+
  scale_fill_grey(start = 0.75, end = 0.25)+
  labs(x = "Treatments",
       y = "Weighted diameter")

ggsave("wd.png")

```



model_WD=dt$WD ~ dt$Treatments
ANOVA=aov(model_WD, data = dt)
Tukey <- TukeyHSD(x=ANOVA, conf.level=0.95)
cld <- multcompLetters4(ANOVA, Tukey)
cld_tukey <- function(Tukey){
   letters_tukey <- multcompLetters(Tukey$`Treatments`)
  return(letters$letters_tukey)
}
tukey_letters <- cld_tukey(Tukey)

## SOC fraction of 2mm&0.25mm
```{r}
dt %>% 
  group_by(Agg_size) %>% 
  summarise(n = n(),
            mean = mean(SOC_fr_g),
            sd = sd(SOC_fr_g),
            se = sd/sqrt(n)
            )

dt %>% 
  group_by(Agg_size, Treatments) %>%
  summarise(n = n(),
            mean = mean(SOC_fr_g),
            sd = sd(SOC_fr_g),
            se = sd/sqrt(n),
            .groups = "drop" # override argument (species & sex)
            )

selected_sizes <- c("2mm", "0.25mm")

dt %>% 
  filter(Agg_size %in% selected_sizes) %>%
  ggplot(., aes(x = Treatments,
                y = SOC_fr_g,
                color = Treatments,
                fill = Treatments,
                )
         ) +
  geom_boxplot(alpha = 0.3)+
  stat_summary(fun = mean, geom = "point", shape = 17, size = 3)+
  facet_wrap(~Agg_size)+
  labs(x = "Treatments",
       y = "Soil Organic Carbon (SOC) in g")

ggsave("soc.png")

#add tukey
#https://r-graph-gallery.com/84-tukey-test.html

```
## MBC fraction of 2mm&0.25mm
```{r}
dt %>% 
  group_by(Agg_size) %>% 
  summarise(n_mbc = n(),
            mean_mbc = mean(MBC_fr_g),
            sd_mbc = sd(SOC_fr_g),
            se_mbc = sd_mbc/sqrt(n_mbc)
            )

dt %>% 
  group_by(Agg_size, Treatments) %>%
  summarise(n_mbc = n(),
            mean_mbc = mean(MBC_fr_g),
            sd_mbc = sd(MBC_fr_g),
            se_mbc = sd_mbc/sqrt(n_mbc),
            .groups = "drop" # override argument (species & sex)
            )

selected_sizes_mbc <- c("2mm", "0.25mm")

dt %>% 
  filter(Agg_size %in% selected_sizes_mbc) %>%
  ggplot(., aes(x = Treatments,
                y = MBC_fr_g,
                color = Treatments,
                fill = Treatments,
                )
         ) +
  geom_boxplot(alpha = 0.3)+
  stat_summary(fun = mean, geom = "point", shape = 17, size = 3)+
  facet_wrap(~Agg_size)+
  labs(x = "Treatments",
       y = "Microbial biomass (MBC) in g")

ggsave("mbc.png")

#add tukey
#https://r-graph-gallery.com/84-tukey-test.html

```

## Train test by 80/20
```{r Train test split data}

# Assuming your data frame is named 'your_data_frame' and you want to remove 'column_to_remove'
#data <- data[, -which(names(data) == "Treatment")]

set.seed(123)
dt <- dt[c(1:2, 4:6)]

train_id <- sample(1:nrow(dt), size = nrow(dt)*0.8)
test_id <- setdiff(1:nrow(dt), train_id)
head(train_id)

# split the train test data
data_train <- dt %>% slice(train_id) 
data_test <- dt %>% slice(test_id)

#slice function is for selecting entire row
#select function is for selecting entire column

#model
model_lm <- lm(WD ~., data_train)
summary(model_lm)


```

## Train data with cross validation 
```{r}
# Example of 5-fold cross-validation
set.seed(123)


# Specify the control parameters for cross-validation
ctrl <- trainControl(method = "cv", number = 5)

# Train a model (e.g., linear regression) using k-fold cross-validation
model_cross <- train(WD ~ ., data = dt, method = "lm", trControl = ctrl)

# View the results
print(model_cross)
summary(model_cross)
```

# Assess R2 values, visualize prediction vs observation and finally variable importance
```{r}
#prediction of test data 
predict_lm        <- predict(model_lm, data_test)
predict_cross     <- predict(model_cross, data_test)

plot(predict_lm, data_test$WD)
plot(predict_cross, data_test$WD)

#R2 for multiple linear model
cor(predict.lm, data_test$WD)^2    #R2 for multiple linear model
cor(predict.m, data_test$WD)^2  

```  
## visualize the prediction vs observation

```{r}
test_result <- data.frame(
  observation = data_test$WD,
  predict_lm = predict_lm,
  predict_cross = predict_cross
  )
#library(GGally)
test_result %>% 
  ggpairs()

ggsave("ml.png")
```  

## visualize the variable importance

```{r}
a <- vip(model_lm) + labs(title = "a. multiple linear model")
b <- vip(model_cross) + labs(title = "b. cross validation model")
a + b

ggsave("var_imp.png")



```
c <- partial(model_lm, pred.var = "Treatments") %>% autoplot() + labs(title = "c. multiple linear model")
d <- partial(model_cross, pred.var = "Treatments") %>% autoplot() + labs(title = "d. cross validation model")

e <- partial(model_lm, pred.var = c("Treatments", "Agg_size")) %>% autoplot() + labs(title = "e. multiple linear model")
f <- partial(model_cross, pred.var = c("Treatments", "Agg|_size")) %>% autoplot() + labs(title = "f. cross validation model")e <- partial(model_lm, pred.var = c("WD_mm", "SOC_g")) %>% autoplot() + labs(title = "e. multiple linear model")


## normalized data
```{r normalized data}
#colMeans(num_data)
norm_data <- (scale(num_data))
#head(norm_data)
```  
## compute the corrleation matrix  
```{r corr matrix}
corr_matrix <- cor(norm_data)
ggcorrplot(corr_matrix)
```  
The higher the value, the most positively correlated the two variables are.
The closer the value to -1, the most negatively correlated they are.  
## applying pca  
```{r}
pca_data <- princomp(corr_matrix)
summary(pca_data)
```
  
Each component explains a percentage of the total variance in the data set. In the Cumulative Proportion section, the first principal component explains almost 77% of the total variance. This implies that almost two-thirds of the data in the set of 9 variables can be represented by just the first principal component. The second one explains 12.08% of the total variance. 

The cumulative proportion of Comp.1 and Comp.2 explains nearly 89% of the total variance. This means that the first two principal components can accurately represent the data. 

It’s great to have the first two components, but what do they really mean? 

This can be answered by exploring how they relate to each column using the loadings of each principal component. 

```{r relation of two component1&2}
pca_data$loadings[, 1:2]
```
  
The loading matrix shows that the first principal component has high positive values for both mwd, soc, Na, Mg, S, Fe_P, MBC. However, the values for P, Fe_DCB are relatively negative. This suggests that soil with a higher aggregation Carbon, Na, Mg, S, Fe_P are excess, while soil with a lower aggregation are in deficit of these minerals.


```{r}
#install.packages("factoextra")
library(factoextra)
fviz_eig(pca_data, addlabels = TRUE)

ggsave("scree.png")
```
  
```{r}
# Graph of the variables
fviz_pca_var(pca_data, col.var = "black")
fviz_pca_var(pca_data, col.var = "cos2",
            gradient.cols = c("#FF004D", "#38E54D", "#4942E4"),
            repel = TRUE)

ggsave("pca.png")

```
fviz_pca_ind(pca_data,
             geom.ind = "point", # show points only (nbut not "text")
             col.ind = data$Treatment, # color by groups
             palette = c("#00AFBB", "#E7B800", "#FC4E07"),
             addEllipses = TRUE, # Concentration ellipses
             legend.title = "Groups"
             )

```{r}
# Create the dataset
data <- data.frame(
  category = rep(c("a", "a", "a", "b", "b", "b"), each = 3),
  fruit = rep(c("apple", "orange", "banana"), times = 2 * 3),  # Corrected replication
  value = c(2, 10, 30, 3, 11, 31, 4, 14, 40, 5, 15, 41)
)

# Rest of the code remains the same
# Step 2: Sum up values for each group
sum_data <- data %>%
  group_by(category, fruit) %>%
  summarize(sum_value = sum(value))

# Step 3: Calculate the average of two 3a and two 3b
average_data <- sum_data %>%
  group_by(category) %>%
  summarize(average_value = mean(sum_value))

# Step 4: Plot the average values
library(ggplot2)

ggplot(average_data, aes(x = category, y = average_value, fill = category)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Values of 3a and 3b", x = "Category", y = "Average Value") +
  theme_minimal()


```
